---
title: "HW_12"
author: "JingjingQi"
date: "4/5/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
data.frame(Feature = c("Pol II presence",
                       "open chromatin vs. repressed",
                       "H3K4me3, H3K27ac, H3K4me1",
                       "TF binding, e.g. p300",
                       "CTCF binding",
                       "chromosomal interactions",
                       "DNA methylation"),
           NGS_Assay = c("detecting transcripts (transcript production is dependent on Pol II!): RNA-seq, PRO-seq, GRO-seq; detecting Pol II itself: ChIP-seq",
                         "detection promoters, gene bodies, enhancers, TFBS: RNA-seq, ATAC-seq, Chip-seq",
                         "enhancers enriched around TSSs: ATAC-seq",
                         "protein DNA binding: Chip-seq",
                         "insulator binding: Chip-seq",
                         "Hi-C, 3C, ChIA-PET",
                         "WGBS, RRBS"))%>%
  kableExtra::kable()%>%kableExtra::kable_styling()
```
###Why is ChIP-seq so much more costly and difficult than ATAC-seq?    
Antibody is needed for enrichment of DNA bound to the protein of interest, needs large number of cells, protocal takes longer all aid up to the cost.  
Antibody binding efficiency specifity might introduce bias.  
Library bias towards shorter fragments, sonication or MNase digestion both might introduce bias.  

###Why is it benefitial to combine information from RNA-seq, ChIP-seq and ATAC-seq experiments? What type of information can you gain from each experiment individidually – and why does it help to integrate them?    
Able to answer refined questions in the same experiment, since chip-seq has higher noise ratio, the other two also helps to proof read. Identify chromatin accessibility, gene expression, and association with regulatory elements. 
