---
title: "HW6"
author: "JingjingQi"
date: "2/12/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Exercises (14 pts)  
###1 Write a script that will: (3pt)  
    run BWA on one of the samples from the Gierlinski dataset  
    run STAR on the same sample  
Remember those three checks after read alignment:  
Is it a BAM file?   STAR: BAM file by default,  BWA: requires convert  
Is it sorted?  STAR: commond option to sort by coordinate, BWA: requires SAMTOOLS to sort by coordinate  
Is it indexed? requires SAMTOOLS to index  

```{eavl = F}
spack load star@2.7.0e
spack load samtools@1.9%gcc@6.3.0
spack load bwa@0.7.15%gcc@6.3.0

$ wget "http://hgdownload.soe.ucsc.edu/goldenPath/sacCer3/bigZips/sacCer3.2bit"
$ wget "http://hgdownload.cse.ucsc.edu/admin/exe/linux.x86_64/twoBitToFa"
$ ../twoBitToFa sacCer3.2bit sacCer3.fa    ## convert 2bit file to Fasta file

#### build index
$ STAR --runMode genomeGenerate \
--runThreadN 4 \
--genomeDir sacCer3_STARindex \
--genomeFastaFiles SacCer3.fa/sacCer3.fasta \   # a file or combined file
--sjdbGTFfile sacCer3.sgd.gtf \
--sjdbOverhang 50  # refer to fastqc file read length distribution 


# pooled file from rimed WT biorep1 option -s 10
$ for file in $(ls | egrep ".*.gz")   
> do
> zcat $file >> WT_biorep1_cmb
> done
```

```{r, eval=F}

$ STAR --runMode alignReads \
--runThreadN 12 \
--genomeDir sacCer3_STARindex \
--readFilesIn /athena/angsd/scratch/jiq4001/HW_Wk4/WT/biorep1/trimed_soft/WT_biorep1_cmb \
--outFileNamePrefix Star_align/WT_biorep1_cmb. \
--outSAMattributes NH HI AS nM MD \
--alignIntronMin 30 \
--alignIntronMax 1000 \ #distribution of yeast intron: http://ribonode.ucsc.edu/Pubs/Spingolaetal99.pdf  
--twopassMode Basic \  #--outSJfilterOverhangMin default: 30 12 12 12
--outSAMtype BAM SortedByCoordinate


$ samtools index WT_biorep1_cmb.Aligned.sortedByCoord.out.bam
```
 

```{r, eval=F}
$ bwa index -p sacCer3_BWAindex/sacCer3 SacCer3.fa/sacCer3.fasta # creat index
$ bwa mem -t 4 -k 20 sacCer3_BWAindex/sacCer3  /athena/angsd/scratch/jiq4001/HW_Wk4/WT/biorep1/trimed_soft/WT_biorep1_cmb > BWA_align/WT_biorep1_cmb.bwa.sam 
$ samtools view -b BWA_align/WT_biorep1_cmb.bwa.sam | samtools sort -o BWA_align/WT_biorep1_cmb.bwa.bam
$ rm WT_biorep1_cmb.bwa.sam
$ samtools index BWA_align/WT_biorep1_cmb.bwa.bam
```

In script: 
```{r, eval=FALSE}
#!/bin/bash
#SBATCH --nodes=1
#SBATCH --ntasks=1 --cpus-per-task=16
#SBATCH --job-name=tmle
#SBATCH --time=7-00:00:00
#SBATCH --mem=80G
#SBATCH --error=/athena/angsd/scratch/jiq4001/error_1
#SBATCH --out=/athena/angsd/scratch/jiq4001/out_1

spack load star@2.7.0e
spack load samtools@1.9%gcc@6.3.0
spack load bwa@0.7.15%gcc@6.3.0

# usage star_bwa_bash <fastq file> <star_dir> <bam_dir>
#check arguments count
arg_count=$#
if [ $arg_count -lt 3]
 then 
  echo "not enough argment <fastq_file> <star_dir> <bam_dir>"
  exist
fi

fastq_file=$1
star_dir=$2
bam_dir=$3

#check if output dir exist
if [ ! -d Star_align ]
 then
  mkdir Star_align
fi

if [ ! -d BWA_align ]
 then
  mkdir BWA_align
fi

STAR --runMode alignReads \
--runThreadN 12 \
--genomeDir ${star_dir} \
--readFilesIn ${fastq_file} \
--outFileNamePrefix Star_align/${fastq_file}. \
--outSAMattributes NH HI AS nM MD \
--alignIntronMin 30 \
--alignIntronMax 1000 \ #distribution of yeast intron: http://ribonode.ucsc.edu/Pubs/Spingolaetal99.pdf  
--twopassMode Basic \  #--outSJfilterOverhangMin default: 30 12 12 12
--outSAMtype BAM SortedByCoordinate
samtools index Star_align/${fastq_file}.Aligned.sortedByCoord.out.bam

bwa mem -t 8 ${bam_dir}  ${fastq_file} > BWA_align/temp 
samtools view -b BWA_align/temp | samtools sort -o BWA_align/${fastq_file}.bwa.bam
rm temp
samtools index BWA_align/${fastq_file}.bwa.bam
```


###2 Subset the aligned reads to select only those that map to chromosome I. (1pt)  
```{r, eval=F}

$ samtools view -h Star_align/WT_biorep1_cmb.Aligned.sortedByCoord.out.bam | egrep "\schrI\s" | wc -l
172054


$ samtools view -h BWA_align/WT_biorep1_cmb.bwa.bam | egrep "\schrI\s" | wc -l
160075

```

###3 Compare the output from BWA and STAR, and summarize any results or differences.  
```{r, eval=F}
STAR
ERR458493.552967  16 chrI 140 255 12M61232N37M2S  *  0  0  CCACTCGTTCACCAGGGCCGGCGGGCTGATCACTTTATCGTGCATCTTGGC     BB?HHJJIGHHJIGIIJJIJGIJIJJIIIGHBJJJJJJHHHHFFDDDA1+B     NH:i:1  HI:i:1  AS:i:41 nM:i:2

BWA
ERR458494.778282  272 chrI  139  0  13H38M  *   0   0       
*   *   NM:i:1  MD:Z:35T2       AS:i:35
```
Star_align  
Number of input reads |	7009665
Uniquely mapped reads % |	85.73%  
Average mapped length |	50.42  
Number of splices: Total |	10688  
Mismatch rate per base, % |	0.35%  
% of reads mapped to multiple loci |	10.21%  


BWA_align  
raw total sequences:	7009665  
reads mapped:	6703899  
average first fragment length:	51  
error rate:	4.692696e-03	# mismatches / bases mapped  
average quality:	36.2  


###Which optional SAM fields does STAR add and what do they represent? (1pt)   
    NH: Number of reported alignments that contain the query in the current record  
    HI: Query hit index  
    AS: Alignment score generated by aligner 
    nM: Edit distance to the reference  
###Which optional SAM fields does BWA add and what do they represent? (1pt)  
    NM: Edit distance to the reference 
    MD: String for mismatching positions  
    AS: Alignment score generated by aligner  
    XS: indicate non multiple mapping
###4 Run bamqc on your BAM files (Note: this is a tool that’s not available in spack, but you can use it via /softlib/apps/EL7/BamQC/bin/bamqc after logging on to a compute node). You will need to figure out how to run this on your own (hint: /softlib/apps/EL7/BamQC/bin/bamqc --help).  
```{r, eval=F}
$ /softlib/apps/EL7/BamQC/bin/bamqc -v
$ BamQC v0.1.25_devel

$ /softlib/apps/EL7/BamQC/bin/bamqc Star_align/WT_biorep1_cmb.Aligned.sortedByCoord.out.bam -f sacCer3.sgd.gtf -g SacCer3.fa/ -o Star_align/ --extract

$ /softlib/apps/EL7/BamQC/bin/bamqc BWA_align/WT_biorep1_cmb.bwa.bam -f sacCer3.sgd.gtf -g SacCer3.fa/ -o BWA_align/ --extract

```
   
###Describe 3 differences between the bamqc results for both the BWA and the STAR output files. (3pt)  
STAR  
Percent primary alignments	88.689  
Percent sequences spliced	0.784  
Percent indels	0.004  

BWA  
Percent primary alignments	99.998
Percent sequences unmapped	4.362  
Percent indels	0.006 (Ins: 0.002; Del: 0.004)  
Percent SNPs	0.443 
  
Percent primary alignments: STAR is lower than BWA  
Percent sequences unmapped: STAR is Lower than BWA  
Percent sequences spliced: STAR non-zero, BWA 0  
Indel and SNP peak show in BWA     

STAR is spice-aware aligner, BWA is not, therefore Star showed lower primary alignment, lower unmapped sequence. and try to finde non matched base with insertion some other locations.  
    
###5 Explain the difference between alignment score and mapping quality in SAM/BAM files. How does the interpretation of the mapping quality field differ between STAR and BWA? (2pt)  
Alignment score quantifies the similarity between read and a reference sequence  
Mapping quality quantifies the probability that a read is correctly aligned to the mapped position.  

###6 What is the difference between a multi-mapping read, and a split read? Find a read that has been split in STAR. How did BWA handle the mapping of that read? (2pt)  
multi-mapping read: a sequence that might be aligned to more than one locations in the genome.  
split read: ead that have two or more alignments to the reference from unique region of the read.  
```{r, eval=FALSE}
[jiq4001@node003 HW_Wk6]$ samtools view -h Star_align/WT_biorep1_cmb.Aligned.sortedByCoord.out.bam | head -n 100 | egrep ".*([0-9]+N)+.*"
# N denotes large insert
```
An example output: ERR458495.840632	0	chrI	12654	3	26M48N25M  

only in Star processed file, BWA doesnt recognize splices.  

###7 How can you remove the unmapped reads from the BWA output? (hint: go back to the notes where FLAG values were explained) (1pt)
```{r, eval=FALSE}
# f output flag, F supress flag    
# unmapped flag code: 4
$ samtools view -b -F 4 BWA_align/WT_biorep1_cmb.bwa.bam > BWA_align/WT_biorep1_cmb.bwa_mapped.bam
```

##Project work (5 pts)  
###If you need a different program than what we have used in the class, you can use spack find to see if the tool is already installed and loadable via spack. If your tool is not there, get in touch with scu@med.cornell.edu to ask them to install it for you.  
If you have processes that will take a long time, go back to the notes from the first day and try to make use of sbatch.  

###1 Download at least one FASTQ file that you will be working with for your project. Document the following details: (2pt)  
  
###where did you get it from?   
    https://www.ncbi.nlm.nih.gov/Traces/study/?acc=SRP026013&o=rna_preparation_sam_s%3Aa%3Bacc_s%3Aa&s=SRR900275,SRR900276,SRR900277,SRR900278,SRR900279,SRR900280,SRR900281,SRR900282,SRR900283,SRR900284,SRR900285,SRR900286,SRR900287,SRR900288,SRR900289,SRR900290  
  
###what publication is it linked to?   
    https://www.ncbi.nlm.nih.gov//pubmed/23977132   
###who generated the data?  
    Mayo Clinic College of Medicine  
###how was the NA extracted?  
    TRIzol kit (Invitrogen)  
###what library prep was used?  
    Illumina PolyA, NuGEN Ovation  
###what cell type was used?  
    human mammary epithelial cell (HMEC) lines obtained from American Type Culture Collection (ATCC)  
###what was the treatment/experimental condition? 
    involuted phenotype associated with an increased breast cancer risk  
###what sequencing platform was used?
    NuGEN---GAIIx sequencer, and Illumina PolyA---HiSeq 2000 at Mayo Clinical  
```{r, eval=FALSE}

#!/bin/bash
#SBATCH --nodes=1
#SBATCH --ntasks=1 --cpus-per-task=16
#SBATCH --job-name=download
#SBATCH --time=7-00:00:00
#SBATCH --mem=80G
#SBATCH --error=/athena/angsd/scratch/jiq4001/error_1
#SBATCH --out=/athena/angsd/scratch/jiq4001/out_1

spack load -r sra-toolkit@2.8.2-1

for a in $(cat /athena/angsd/scratch/jiq4001/Pj/Acc_ilumina.txt)
do
 fastq-dump --split-files $a  --outdir /athena/angsd/scratch/jiq4001/Pj/test --gzip
done
```
    
###2 Align the FASTQ file with an appropriate aligner (you may have to build a new index). Document: (3pt)  
parameters (and why you chose them)  
summary of outcome and basic QC  
###Record:  
Genome:  
http://hgdownload.soe.ucsc.edu/goldenPath/hg38/bigZips/hg38.2bit  
Annotation:  
http://hgdownload.soe.ucsc.edu/goldenPath/hg38/bigZips/genes/  
hg38.refGene.gtf.gz        10-Jan-2020 09:33   23M  
genomeGenerate: --sjdbOverhang 99  
alignReads: --alignIntronMin 20 --alignIntronMax 1000000 --twopassMode Basic  

```{r, eval=FALSE}
$ wget "http://hgdownload.soe.ucsc.edu/goldenPath/hg38/bigZips/hg38.2bit"
$/athena/angsd/scratch/jiq4001/HW_Wk6/twoBitToFa hg38.2bit hg38.fa

$ wget "http://hgdownload.soe.ucsc.edu/goldenPath/hg38/bigZips/genes/hg38.refGene.gtf.gz "

#!/bin/bash
#SBATCH --nodes=1
#SBATCH --ntasks=1 --cpus-per-task=16
#SBATCH --job-name=tmle
#SBATCH --time=7-00:00:00
#SBATCH --mem=80G
#SBATCH --error=/athena/angsd/scratch/jiq4001/error

spack load star@2.7.0e

cwd=/athena/angsd/scratch/jiq4001/Pj
  
STAR --runMode genomeGenerate \
--runThreadN 16 \
--genomeDir /athena/angsd/scratch/jiq4001/Pj/hg38_STARindex   \
--genomeFastaFiles /athena/angsd/scratch/jiq4001/Pj/hg38.fa   \
--sjdbGTFfile /athena/angsd/scratch/jiq4001/Pj/hg38.ensGene.gtf  \
--sjdbOverhang 99 # refered by sample fastqc read length  
```

```{r,eval=FALSE}
trim_galore --illumina --dont_gzip --stringency 5 --length 30 --paired -o trimed/ test/run_test/SRR900275_1.fastq.gz test/run_test/SRR900275_2.fastq.gz
```

```{r, eval=FALSE}
#!/bin/bash
#SBATCH --nodes=4
#SBATCH --ntasks=1 --cpus-per-task=16
#SBATCH --job-name=tmle
#SBATCH --time=7-00:00:00
#SBATCH --mem=80G
#SBATCH --error=/athena/angsd/scratch/jiq4001/error 
spack load fastqc
spack load -r trimgalore
spack load -r py-multiqc
# set current work dir
cwd=/athena/angsd/scratch/jiq4001/Pj

mkdir ${cwd}/QC
mkdir ${cwd}/trimed
mkdir ${cwd}/QC_trimed
mkdir ${cwd}/Star_align
mkdir ${cwd}/multi_qc

input_dir=${cwd}/test
qc_dir=${cwd}/QC
trim_dir=${cwd}/trimed
qc_trim_dir=${cwd}/QC_trimed
multi_qc=${cwd}/multi_qc

for file in $(ls ${input_dir} | cut -b -9 | uniq -d)
do
    fastqc ${input_dir}/${file}_1.fastq.gz -o ${qc_dir} -t 16 --extract
    fastqc ${input_dir}/${file}_2.fastq.gz -o ${qc_dir} -t 16 --extract
    
    trim_galore --illumina --stringency 5 --length 30 --paired -o ${trim_dir} ${input_dir}/${file}_1.fastq.gz ${input_dir}/${file}_2.fastq.gz
    
    fastqc ${trim_dir}/${file}_1_val_1.fq.gz -o ${qc_trim_dir} -t 16 --extract
    fastqc ${trim_dir}/${file}_2_val_2.fq.gz -o ${qc_trim_dir} -t 16 --extract
    
done


multiqc ${qc_dir}
multiqc ${qc_trim_dir}
```



```{r, eval=FALSE}
#!/bin/bash
#SBATCH --nodes=1
#SBATCH --ntasks=1 --cpus-per-task=16
#SBATCH --job-name=tmle
#SBATCH --time=7-00:00:00
#SBATCH --mem=80G
#SBATCH --error=/athena/angsd/scratch/jiq4001/error 
cwd=/athena/angsd/scratch/jiq4001/Pj

mkdir ${cwd}/Star_align
trim_dir=${cwd}/trimed
star_align=${cwd}/Star_align

for file in $(ls ${trim_dir} | cut -b -9 | uniq -d)
do
STAR --runMode alignReads \
--runThreadN 16 \
--genomeDir ${cwd}/hg38_STARindex \
--readFilesIn ${trim_dir}/${file}_1_val_1.fq.gz, ${trim_dir}/${file}_2_val_2.fq.gz \
--outFileNamePrefix ${star_align}/${file}. \
--outSAMattributes NH HI AS nM MD \
--twopassMode Basic \
--readFilesCommand zcat \
--outSAMtype BAM SortedByCoordinate 
done

# ENCODE options default for human 
#About 5.24% of introns are more than 200,000 bp and less than 10% of introns are more than 11,000 bp in length. Also, < 0.01% of the introns are < 20 bp in length   --M.K. Sakharkar et al. / Distributions of Exons and Introns in the Human Genome
```
                      Number of input reads |	40909112  
                      Average input read length |	49  
                                    UNIQUE READS:  
                   Uniquely mapped reads number |	33754010  
                        Uniquely mapped reads % |	82.51%  
                          Average mapped length |	49.57  
                       Number of splices: Total |	5786990  
            Number of splices: Annotated (sjdb) |	5777530  
                       Number of splices: GT/AG |	5735122  
                       Number of splices: GC/AG |	39239  
                       Number of splices: AT/AC |	3807  
               Number of splices: Non-canonical |	8822  
                      Mismatch rate per base, % |	0.13%  
                         Deletion rate per base |	0.00%  
                        Deletion average length |	1.51  
                        Insertion rate per base |	0.00%  
                       Insertion average length |	1.28  
                             MULTI-MAPPING READS:         
        Number of reads mapped to multiple loci |	6593238  
             % of reads mapped to multiple loci |	16.12%  
        Number of reads mapped to too many loci |	178844  
             % of reads mapped to too many loci |	0.44%  
                                  UNMAPPED READS:  
       % of reads unmapped: too many mismatches |	0.00%  
                 % of reads unmapped: too short |	0.73%  
                     % of reads unmapped: other |	0.20%  
                                  CHIMERIC READS:   
                       Number of chimeric reads |	0  
                            % of chimeric reads |	0.00%  

```{r, eval=FALSE}
https://bmcbioinformatics.biomedcentral.com/articles/10.1186/1471-2105-14-S11-S8 # impack of annotation choice  
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4631051/
```

**NOTE: need to find out NuGEN Ovation kit's adapter sequence**
