---
title: "HW6"
author: "JingjingQi"
date: "2/12/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Exercises (14 pts)  
###1 Write a script that will: (3pt)  
    run BWA on one of the samples from the Gierlinski dataset  
    run STAR on the same sample  
Remember those three checks after read alignment:  
Is it a BAM file?   STAR: BAM file by default,  BWA: requires convert  
Is it sorted?  STAR: commond option to sort by coordinate, BWA: requires SAMTOOLS to sort by coordinate  
Is it indexed?  STAR: indexed,   BWA: requires SAMTOOLS to index  
```{eavl = F}
spack load star@2.7.0e
spack load samtools@1.9%gcc@6.3.0
spack load bwa@0.7.15%gcc@6.3.0

wget "http://hgdownload.soe.ucsc.edu/goldenPath/sacCer3/bigZips/sacCer3.2bit"
wget "http://hgdownload.cse.ucsc.edu/admin/exe/linux.x86_64/twoBitToFa"
[jiq4001@node003 sacCer3.fa]$../twoBitToFa sacCer3.2bit sacCer3.fa    ## convert 2bit file to Fasta file


STAR --runMode genomeGenerate \
--runThreadN 4 \
--genomeDir sacCer3_STARindex \
--genomeFastaFiles SacCer3.fa/sacCer3.fasta \   # a file or combined file
--sjdbGTFfile sacCer3.sgd.gtf \
--sjdbOverhang 50  # refer to fastqc file read length distribution 

```

```{r, eval=F}
# pooled file from rimed WT biorep1 option -s 10
[jiq4001@node003 trimed_soft]$ for file in $(ls | egrep ".*.gz")   
> do
> zcat $file >> WT_biorep1_cmb
> done

STAR --runMode alignReads \
--runThreadN 12 \
--genomeDir sacCer3_STARindex \
--readFilesIn /athena/angsd/scratch/jiq4001/HW_Wk4/WT/biorep1/trimed_soft/WT_biorep1_cmb \
--outFileNamePrefix Star_align/WT_biorep1_cmb. \
--outSAMattributes NH HI AS nM MD \
--alignIntronMin 30 \
--alignIntronMax 1000 \ #distribution of yeast intron: http://ribonode.ucsc.edu/Pubs/Spingolaetal99.pdf  
--twopassMode Basic \  #--outSJfilterOverhangMin default: 30 12 12 12
--outSAMtype BAM SortedByCoordinate


[jiq4001@node003 Star_align]$ samtools index WT_biorep1_cmb.Aligned.sortedByCoord.out.bam
```
 


```{r, eval=F}
[jiq4001@node003 HW_Wk6]$ bwa index -p sacCer3_BWAindex/sacCer3 SacCer3.fa/sacCer3.fasta # creat index
[jiq4001@node003 HW_Wk6]$ bwa mem -t 4 -k 20 sacCer3_BWAindex/sacCer3  /athena/angsd/scratch/jiq4001/HW_Wk4/WT/biorep1/trimed_soft/WT_biorep1_cmb > BWA_align/WT_biorep1_cmb.bwa.sam 
[jiq4001@node003 HW_Wk6]$ samtools view -b BWA_align/WT_biorep1_cmb.bwa.sam | samtools sort -o BWA_align/WT_biorep1_cmb.bwa.bam
[jiq4001@node003 HW_Wk6]$ rm WT_biorep1_cmb.bwa.sam

```


###2 Subset the aligned reads to select only those that map to chromosome I. (1pt)  
```{r, eval=F}

[jiq4001@node003 Star_align]$ samtools view -h Star_align/WT_biorep1_cmb.Aligned.sortedByCoord.out.bam | egrep "\schrI\s" | wc -l
172054


[jiq4001@node003 HW_Wk6]$ samtools view -h BWA_align/WT_biorep1_cmb.bwa.bam | egrep "\schrI\s" | wc -l
160075

```

###3 Compare the output from BWA and STAR, and summarize any results or differences.  
```{r, eval=F}
STAR
ERR458493.552967  16 chrI 140 255 12M61232N37M2S  *  0  0  CCACTCGTTCACCAGGGCCGGCGGGCTGATCACTTTATCGTGCATCTTGGC     BB?HHJJIGHHJIGIIJJIJGIJIJJIIIGHBJJJJJJHHHHFFDDDA1+B     NH:i:1  HI:i:1  AS:i:41 nM:i:2

BWA
ERR458494.778282  272 chrI  139  0  13H38M  *   0   0       
*   *   NM:i:1  MD:Z:35T2       AS:i:35
```
    ###Which optional SAM fields does STAR add and what do they represent? (1pt)   
    NH: Number of reported alignments that contain the query in the current record  
    HI: Query hit index  
    AS: Alignment score generated by aligner 
    nM: Edit distance to the reference  
    ###Which optional SAM fields does BWA add and what do they represent? (1pt)  
    NM: Edit distance to the reference 
    MD: String for mismatching positions  
    AS: Alignment score generated by aligner  
    XS: indicate non multiple mapping
###4 Run bamqc on your BAM files (Note: this is a tool that’s not available in spack, but you can use it via /softlib/apps/EL7/BamQC/bin/bamqc after logging on to a compute node). You will need to figure out how to run this on your own (hint: /softlib/apps/EL7/BamQC/bin/bamqc --help).  
```{r, eval=F}
[jiq4001@node003 HW_Wk6]$ /softlib/apps/EL7/BamQC/bin/bamqc -v
[jiq4001@node003 HW_Wk6]$ BamQC v0.1.25_devel

[jiq4001@node003 HW_Wk6]$ /softlib/apps/EL7/BamQC/bin/bamqc Star_align/WT_biorep1_cmb.Aligned.sortedByCoord.out.bam -f sacCer3.sgd.gtf -g SacCer3.fa/ -o Star_align/ --extract

[jiq4001@node003 HW_Wk6]$ /softlib/apps/EL7/BamQC/bin/bamqc BWA_align/WT_biorep1_cmb.bwa.bam -f sacCer3.sgd.gtf -g SacCer3.fa/ -o BWA_align/ --extract

```
    ###Describe 3 differences between the bamqc results for both the BWA and the STAR output files. (3pt)  
    Percent primary alignments: STAR is lower than BWA  
    Percent sequences unmapped: STAR is Lower than BWA  
    Percent sequences spliced: STAR non-zero, BWA 0  
    STAR is spice-aware aligner, BWA is not, therefore Star showed lower primary alignment, lower unmapped sequence.  
    
###5 Explain the difference between alignment score and mapping quality in SAM/BAM files. How does the interpretation of the mapping quality field differ between STAR and BWA? (2pt)  
Alignment score quantifies the similarity between read and a reference sequence  
Mapping quality quantifies the probability that a read is correctly aligned to the mapped position.  

###6 What is the difference between a multi-mapping read, and a split read? Find a read that has been split in STAR. How did BWA handle the mapping of that read? (2pt)  
multi-mapping read: a sequence that might be aligned to more than one locations in the genome.  
split read: ead that have two or more alignments to the reference from unique region of the read.  
```{r, eval=FALSE}
[jiq4001@node003 HW_Wk6]$ samtools view -h Star_align/WT_biorep1_cmb.Aligned.sortedByCoord.out.bam | head -n 100 | egrep ".*([0-9]+N)+.*"
# N denotes large insert
```
only in Star processed file, BWA doesnt recognize splices.  

###7 How can you remove the unmapped reads from the BWA output? (hint: go back to the notes where FLAG values were explained) (1pt)
```{r, eval=FALSE}
# f output flag, F supress flag    
# unmapped flag code: 4
[jiq4001@node003 HW_Wk6]$ samtools view -b -F 4 BWA_align/WT_biorep1_cmb.bwa.bam > BWA_align/WT_biorep1_cmb.bwa_mapped.bam
```

##Project work (5 pts)  
###If you need a different program than what we have used in the class, you can use spack find to see if the tool is already installed and loadable via spack. If your tool is not there, get in touch with scu@med.cornell.edu to ask them to install it for you.  
If you have processes that will take a long time, go back to the notes from the first day and try to make use of sbatch.  
```{r, eval=FALSE}
spack load -r sra-toolkit@2.8.2-1
```

###1 Download at least one FASTQ file that you will be working with for your project. Document the following details: (2pt)  
    ###where did you get it from?   
    https://www.ncbi.nlm.nih.gov/Traces/study/?acc=SRP026013&o=rna_preparation_sam_s%3Aa%3Bacc_s%3Aa&s=SRR900275,SRR900276,SRR900277,SRR900278,SRR900279,SRR900280,SRR900281,SRR900282,SRR900283,SRR900284,SRR900285,SRR900286,SRR900287,SRR900288,SRR900289,SRR900290  
    ###what publication is it linked to?   
    https://www.ncbi.nlm.nih.gov//pubmed/23977132   
    ###who generated the data?  
    Mayo Clinic College of Medicine  
    ###how was the NA extracted?  
    TRIzol kit (Invitrogen)  
    ###what library prep was used?  
    Illumina PolyA, NuGEN Ovation  
    ###what cell type was used?  
    human mammary epithelial cell (HMEC) lines obtained from American Type Culture Collection (ATCC)  
    ###what was the treatment/experimental condition? 
    involuted phenotype associated with an increased breast cancer risk  
    ###what sequencing platform was used?
    NuGEN---GAIIx sequencer, and Illumina PolyA---HiSeq 2000 at Mayo Clinical  
```{r, eval=FALSE}

#!/bin/bash
#SBATCH --nodes=1
#SBATCH --ntasks=1 --cpus-per-task=16
#SBATCH --job-name=tmle
#SBATCH --time=7-00:00:00
#SBATCH --mem=80G
#SBATCH --error=/athena/angsd/scratch/jiq4001/error_1
#SBATCH --out=/athena/angsd/scratch/jiq4001/out_1

for a in $(cat /athena/angsd/scratch/jiq4001/Pj/Acc_Nugen.txt)
do
 fastq-dump $a --outdir /athena/angsd/scratch/jiq4001/Pj/test --gzip
done
```
    
###2 Align the FASTQ file with an appropriate aligner (you may have to build a new index). Document: (3pt)  
    parameters (and why you chose them)  
    summary of outcome and basic QC  
###Record:  
Genome:  
http://hgdownload.soe.ucsc.edu/goldenPath/hg38/bigZips/hg38.2bit  
Annotation:  
http://hgdownload.soe.ucsc.edu/goldenPath/hg38/bigZips/genes/  
hg38.ensGene.gtf.gz       10-Jan-2020 09:33   27M  
genomeGenerate: --sjdbOverhang 99  
alignReads: --alignIntronMin 20 --alignIntronMax 1000000 --twopassMode Basic  

```{r, eval=FALSE}
[jiq4001@node003 Pj]$ wget "http://hgdownload.soe.ucsc.edu/goldenPath/hg38/bigZips/hg38.2bit"
[jiq4001@node003 Pj]$/athena/angsd/scratch/jiq4001/HW_Wk6/twoBitToFa hg38.2bit hg38.fa

[jiq4001@node003 Pj]$ wget "http://hgdownload.soe.ucsc.edu/goldenPath/hg38/bigZips/genes/hg38.ensGene.gtf.gz"

#!/bin/bash
#SBATCH --nodes=1
#SBATCH --ntasks=1 --cpus-per-task=16
#SBATCH --job-name=tmle
#SBATCH --time=7-00:00:00
#SBATCH --mem=80G
#SBATCH --error=/athena/angsd/scratch/jiq4001/error

spack load star@2.7.0e

cwd=/athena/angsd/scratch/jiq4001/Pj
  
STAR --runMode genomeGenerate \
--runThreadN 16 \
--genomeDir /athena/angsd/scratch/jiq4001/Pj/hg38_STARindex   \
--genomeFastaFiles /athena/angsd/scratch/jiq4001/Pj/hg38.fa   \
--sjdbGTFfile /athena/angsd/scratch/jiq4001/Pj/hg38.ensGene.gtf  \
--sjdbOverhang 99 # refered by sample fastqc read length  
```

```{r, eval=FALSE}
#!/bin/bash
#SBATCH --nodes=4
#SBATCH --ntasks=1 --cpus-per-task=16
#SBATCH --job-name=tmle
#SBATCH --time=7-00:00:00
#SBATCH --mem=80G
#SBATCH --error=/athena/angsd/scratch/jiq4001/error 
spack load fastqc

# set current work dir
cwd=/athena/angsd/scratch/jiq4001/Pj
mkdir ${cwd}/QC


for file in $(ls ${cwd}/test)
  do
    fastqc $file -o ${cwd}/QC/$file -t 4 --extract
done

spack load -r py-multiqc

```

```{r, eval=FALSE}
cwd=/athena/angsd/scratch/jiq4001/Pj
mkdir ${cwd}/Star_align

for file in $(cat ls${cwd}/test)
do
 STAR --runMode alignReads --runThreadN 16 --genomeDir ${cwd}/hg38_STARindex --readFilesIn ${cwd}/test/run_test --outFileNamePrefix ${cwd}/Star_align/${file}. --outSAMattributes NH HI AS nM MD --alignSJDBoverhangMin 20 --alignIntronMin 20  --alignIntronMax 1000000 \  #default #--outSJfilterOverhangMin default: 30 12 12 12
--twopassMode Basic \
--readFilesCommand zcat \
--outSAMtype BAM SortedByCoordinate
 done
 
#About 5.24% of introns are more than 200,000 bp and less than 10% of introns are more than 11,000 bp in length. Also, < 0.01% of the introns are < 20 bp in length   --M.K. Sakharkar et al. / Distributions of Exons and Introns in the Human Genome
```

```{r, eval=FALSE}
https://bmcbioinformatics.biomedcentral.com/articles/10.1186/1471-2105-14-S11-S8 # impack of annotation choice  
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4631051/
```

